RNApyro

### 2012-10-03 Recomb/methods_RECOMB.tex
first draft inside

%!TEX root = main_RECOMB.tex
\section{Methods}
\label{sec:methods}
Let $\Omega$ be an un-gapped RNA alignment, $S$ its associate secondary structure, $s$ an RNA 
sequence and $m\geq 0$. $S$ is considered as  one derivation of the SCFG generating  
all possible secondary structures of the same length as $s$. We define a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990} with the two following functions,
$\Z{i,j}{m}{a,b}$ and $\Y{i,j}{m}{a,b}$. The former, ass the 
equivalent of the \emph{inside}, computes the partition function between $i$ and $j$ included, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$) and containing $m$ mutations. The latter computes the \emph{outside}, in particular, knowing 
that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$) and containing, outside 
including $i$ and $j$, $m$ mutations.
The Boltzmann weights are a combination of the base pairs stacking energy, 
using as values those of the NNDB~\cite{Turner2010}, and
the average isostericity with $\Omega$, as defined in~\cite{Stombaugh2009}.

@@

\subsection{Energy Model}
The energy we will be composed of two function, $ES^{\beta}_{ab\to a'b'}$ and 
$EI_{ab\to a'b'}$. $ES^\beta_{ab\to a'b'}$ is equal to the 
stacking energy of the base pair with nucleotides $ab$ on top of the base pair with nucleotides 
$a'b'$, as set in the NNDB~\cite{Turner2010}. If one of the base pair is not valid (i.e. not in 
$\{\text{GU},\text{UG},\text{CG},\text{GC}, \text{AU or UA}\}$, the value is a parameter 
$\beta \in [1,\infty]$. This allows
to completely forbid a sequence where a base pair is non valid, when $\beta = \infty$ or only 
penalize it.
$EI_{ab\to a'b'}$ is the isostericity between
 the base pair $ab$ and $a'b'$ as defined in~\cite{Stombaugh2009}. Let be $\alpha\in[0,1]$, it 
 will be used to balance the weight given to the stacking energy or the isostericity.

\subsection{Inside}
We define the \emph{Inside} function $\Z{i,j}{m}{a,b}$ as follows.
\subsubsection{Initial conditions}
$$
	\forall i \in (0,\cdots,n-1):\, \Z{i+1,i}{m}{a,b}=\left\{
	\begin{array}{ll}
		1 &\text{If } m = 0\\
		0 &\text{Else }
	\end{array}\right.
$$
\subsubsection{Recursion}
$$
	\Z{i,j}{m}{a,b}:=\left\{
  \begin{array}{ll}
  		\displaystyle
      \sum_{\substack{a'\in \B,\\ \Kron_{a',s[i]}\le m}}  
      \Z{i+1,j}{m-\Kron_{a',s[i]}}{a',b} & \text{If }S{[i]}=-1\\
      \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s[i]s[j]}\le m}}  
			 e^{\frac{-ES_{a b \to a' b'}}{RT}}
			 \Z{i+1,j-1}{m-\Kron_{a'b',s[i]s[j]}}{a',b'}&
			 \text{Elif }S{[i]}=j \land S{[i-1]}=j+1\\
			 \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s[i]s[k]}\le m}}
      \sum_{m'=0}^{m-\Kron_{a'b',s[i]s[k]}}
   		 e^{\frac{-EI_{s[i]s[k],a'b'}}{RT}}
      \Z{i+1,k-1}{m-\Kron_{a'b',s[i]s[k]}-m'}{a',b'}
      \Z{k+1,j}{m'}{b',b} & \text{Elif }S[i]=k \land i < k \leq j\\
      0 &\text{Else}
	\end{array}\right.
$$

\subsection{Outside}	
\subsubsection{Initial conditions}
$$
	\Y{-1,j}{m}{X,X}:=
		\displaystyle
	  \Z{j,n-1}{m}{X,X}
$$
\subsubsection{Recursion}
$$
	\Y{i,j}{m}{a,b} = \left\{
  \begin{array}{ll}
		\displaystyle
    \sum_{\substack{a'\in \B,\\ \Kron_{a',s[i]}\le m}}
    \Y{i-1,j}{m- \Kron_{a',s[i]}}{a',b} &
    \text{Elif }S[i]=-1 \\
    \displaystyle
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s[i]s[j]}\le m}}
		 e^{\frac{-ES_{ a' b' \to a b }}{RT}}
    \Y{i-1,j+1}{m- \Kron_{a'b',s[i]s[j]}}{a',b'} &
   	 \text{Elif }S{[i]}=j \land S{[i+1]}=j-1\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s[i]s[k]}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s[i]s[k]}}
  		 e^{\frac{-EI_{s[i]s[k],a'b'}}{RT}}
		 \Y{i-1,k+1}{m- \Kron_{a'b',s[i]s[k]} - m'}{a',b'}
     \Z{j,k-1}{m'}{b,b'} &
		 \text{Elif }S{[i]}=k \geq j\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s[k]s[i]}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s[k]s[i]}}
  		 e^{\frac{-EI_{s[i]s[k],a'b'}}{RT}}
		 \Y{k-1,j}{m- \Kron_{a'b',s[k]s[i]} - m'}{a',b}
     \Z{k+1,i-1}{m'}{a',b'} &
		 \text{Elif }-1 < S{[i]}=k < i\\
		 0 & \text{Else}
  \end{array}\right.
$$
\subsection*{}
We must have (if $S[k] = -1)$:
$$
	\Z{0,n-1}{m}{X,X} \equiv     
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a};\qquad
	\forall k \in \{0,\cdots,n-1\}
$$

>>>>

%!TEX root = main_RECOMB.tex
\section{Methods}
\label{sec:methods}
Let $\Omega$ be an un-gapped RNA alignment, $S$ its associate secondary structure, $s$ an RNA 
sequence and $m\geq 0$. $S$ is considered as  one derivation of the SCFG generating  
all possible secondary structures of the same length as $s$. We define a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990} with the two following functions,
$\Z{i,j}{m}{a,b}$ and $\Y{i,j}{m}{a,b}$. The former, as the 
equivalent of the \emph{inside}, computes the partition function between $i$ and $j$ included, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$) and containing $m$ mutations. The latter computes the \emph{outside}, in particular, knowing 
that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$) and containing outside,
including $i$ and $j$, $m$ mutations.
The Boltzmann weights are a combination of the base pairs stacking energy, 
using as values those of the NNDB~\cite{Turner2010}, and
the average isostericity with $\Omega$, as defined in~\cite{Stombaugh2009}.

@@

\subsection{Energy Model}
The energy we will be composed of two function, $ES^{\beta}_{ab\to a'b'}$ and 
$EI_{(i,j),ab}$. The former is equal to the 
stacking energy of the base pair with nucleotides $ab$ on top of the base pair with nucleotides 
$a'b'$, as set in the NNDB~\cite{Turner2010}. If one of the base pair is not valid (i.e. not in 
$\{\text{GU},\text{UG},\text{CG},\text{GC}, \text{AU or UA}\}$, the value is a parameter 
$\beta \in [1,\infty]$. This allows
to completely forbid a sequence where a base pair is non valid, when $\beta = \infty$ or only 
penalize it.
$EI_{(i,j), a'b'}$ is the average of the sum of differences between the isostericity
of base pairs at positions $(i,j)$ in $\Omega$ and $s_is_j$, and the isostericity of base pairs
 at positions $(i,j)$ in $\Omega$ and $ab$. If gives us an indication 
 if the base pair $ab$ is more isosteric to the set $\Omega$ than the one on the sequence 
 $\omega$. Where the isostericity values 
 between any pair of canonical base pairs is  as defined in~\cite{Stombaugh2009}. 
 Let be $\alpha\in[0,1]$, it 
 will be used to balance the weight given to the stacking energy or the isostericity.

\subsection{Inside}
To define the \emph{Inside} function $\Z{i,j}{m}{a,b}$ as a recurrence, we will use as initial conditions:

\[
	\forall i \in (0,\cdots,n-1):\, \Z{i+1,i}{m}{a,b}=\left\{
	\begin{array}{ll}
		1 &\text{If } m = 0\\
		0 &\text{Else }
	\end{array}\right.
\]
In other words, when we evaluate the function $\mathcal Z$  over only one nucleotide, there 
is only one possible solution if there is $0$ mutations left, and none else. Since the 
energetics term only consider base pairs, they are not involved in the initial conditions.

The recurrence is composed of four terms:
$$
	\Z{i,j}{m}{a,b}:=\left\{
  \begin{array}{ll}
  		\displaystyle
      \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}  
      \Z{i+1,j}{m-\Kron_{a',s_i}}{a',b} & \text{If }S_{i}=-1\\
      \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
			 e^{\frac{-(\alpha ES_{a b \to a' b'}+(1-\alpha)EI_{(i,j),a'b'})}{RT}}
			 \Z{i+1,j-1}{m-\Kron_{a'b',s_is_j}}{a',b'}&
			 \text{Elif }S_i=j \land S_{i-1}=j+1\\
			 \displaystyle
      \sum_{\substack{a',b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
      \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
   		 e^{\frac{-(1-\alpha)EI_{(i,k),a'b'}}{RT}}
      \Z{i+1,k-1}{m-\Kron_{a'b',s_is_k}-m'}{a',b'}
      \Z{k+1,j}{m'}{b',b} & \text{Elif }S_i=k \land i < k \leq j\\
      0 &\text{Else}
	\end{array}\right.
$$
The cases can be breaked down as follows.
\begin{description}
\item[$S_{i}=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
as if we increase the lower interval bound by $1$ (i.e. $i+1$), and consider all possible
 nucleotides $a'$ at position $i$, taking into account the 
\item[$S_i=j$ and $S_{i-1}=j+1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i-1$ is
paired with $j+1$, we are in the only case were stacked base pairs can occur. We thus add
the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
to compute is the \emph{inside} value of the interval $[i+1,j-1]$ over all possible nucleotides 
$a',b'\in B^2$ at positions $i$ and $j$ respectively.
\item[$S_i=k$ and $i<k \leq j$:] If nucleotide $i$ is paired with position $k$ but is not stacked outside, the 
only term contributing to the energy is the isostericity of the base pair $(i,k)$. This creates
two different intervals for which we must compute the values, $[i+1,k-1]$ and $[k+1,j-1]$, for 
all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $j$ respectively.
\item[Else:] In all other cases, we are in a derivation of the SCFG that does not correspond to the 
secondary structure $S$, and we return $0$.

\end{description}

\subsection{Outside}	
The \emph{Outside} function, $\Y{i,j}{m}{a,b}$, will also be defined as a recurrence and have 
as initial conditions the following.
$$
	\Y{-1,j}{m}{X,X}:=
		\displaystyle
	  \Z{j,n-1}{m}{X,X}
$$
\subsubsection{Recursion}
$$
	\Y{i,j}{m}{a,b} = \left\{
  \begin{array}{ll}
		\displaystyle
    \sum_{\substack{a'\in \B,\\ \Kron_{a',s[i]}\le m}}
    \Y{i-1,j}{m- \Kron_{a',s[i]}}{a',b} &
    \text{Elif }S[i]=-1 \\
    \displaystyle
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s[i]s[j]}\le m}}
		 e^{\frac{-ES_{ a' b' \to a b }}{RT}}
    \Y{i-1,j+1}{m- \Kron_{a'b',s[i]s[j]}}{a',b'} &
   	 \text{Elif }S{[i]}=j \land S{[i+1]}=j-1\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s[i]s[k]}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s[i]s[k]}}
  		 e^{\frac{-EI_{s[i]s[k],a'b'}}{RT}}
		 \Y{i-1,k+1}{m- \Kron_{a'b',s[i]s[k]} - m'}{a',b'}
     \Z{j,k-1}{m'}{b,b'} &
		 \text{Elif }S{[i]}=k \geq j\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s[k]s[i]}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s[k]s[i]}}
  		 e^{\frac{-EI_{s[i]s[k],a'b'}}{RT}}
		 \Y{k-1,j}{m- \Kron_{a'b',s[k]s[i]} - m'}{a',b}
     \Z{k+1,i-1}{m'}{a',b'} &
		 \text{Elif }-1 < S{[i]}=k < i\\
		 0 & \text{Else}
  \end{array}\right.
$$
\subsection*{}
We must have (if $S[k] = -1)$:
$$
	\Z{0,n-1}{m}{X,X} \equiv     
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a};\qquad
	\forall k \in \{0,\cdots,n-1\}
$$
